Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
work on intro

  * why model finding?
  * but SMT solvers better at proving
    * a single universal quantifier in an axiom is enough to put it off track
      (``maybe unsat'')

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13,reynolds-et-al-2013}
have focused on finding models for quantified formulas in SMT.
Ge and de Moura \cite{GeDeM-CAV-09} presented a complete instantiation-based
procedure for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all
occurrences of variables occur as direct subterms of uninterpreted
functions. For example, the formula $\forall x%\typ{\sortint} --- the sort is not important, right?
.\;\, \con{f}( x )
\teq \con{g}( x ) + 5$ is in the essentially uninterpreted fragment.
% since all
% occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and
% $\con{g}$.
Other syntactic criteria are identified in this paper that extend
this fragment slightly, including cases when variables occur as arguments of
arithmetic predicates. The techniques in~Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} can be used for finding finite
models for quantified
formulas over uninterpreted types as well as types having a fixed finite
interpretation, such as fixed-width bit vectors and enumerated datatypes.
Finite model finding techniques can
find a model for the formula $\forall x,\, y : \tau.\;\, x \teq
y \vee \neg \con{f}( x ) \teq \con{f}( y )$ where the uninterpreted type $\tau$ is
interpreted as a finite set and $\con{f}$ as injective finite map.

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions.
In particular, the fragments in~\cite{GeDeM-CAV-09} are not applicable when the argument of a recursive function is used in various ways (e.g. within an arithmetic expression),
while the fragment in~\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} is not applicable for functions over infinite domains such as the integers.
So, for instance, the quantified formula $\forall x : {\sortint}.\;\, \lite\bigl( p \leq 0, p( x ) \teq 0, p( x ) \teq 2 * p( x - 1 ) \bigr)$
is not in the essentially uninterpreted fragment (since $x$ occurs as a child of $-$),
nor does finite model finding apply (since $x$ is of type $\sortint$).

@@

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by
$\Sigma$-interpretation $\J$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, we know $\varphi$ is of the form
$\DDD \wedge \varphi_0$. We first argue that $\I$ satisfies a $\Sigma$-formula
$\psi$ that is closed under function expansion with respect to $\DDD$, and
where $\psi$ entails $\varphi_0$.

>>>>

Many SMT solvers can reason about quantified formulas using incomplete
instantiation-based methods \cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These methods work well in the context of proving (i.e., showing
unsatisfiability), but they generally cannot be used to find models (i.e.,
showing satisfiability). Typically, a single universal quantifier in one of the
axioms of a problem is enough to prevent the discovery of models.

In the past few years, techniques have been developed to finding models for
quantified formulas in SMT.
Ge and de Moura \cite{GeDeM-CAV-09} introduced a complete instantiation-based
procedure for formulas in the \relax{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all
occurrences of variables occur as direct subterms of uninterpreted
functions---e.g., $\forall x%\typ{\sortint} --- the sort is not important, right?
.\;\, \con{f}( x )
\teq \con{g}( x ) + 5$.
% since all
% occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and
% $\con{g}$.
(Other syntactic criteria extend
this fragment slightly, including cases when variables occur as arguments of
arithmetic predicates.) Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} introduced techniques for finding finite
models for quantified
formulas over uninterpreted types as well as types having a fixed finite
interpretation, such as fixed-width bit vectors and enumerated datatypes.
These techniques can
find a model for the formula $\forall x,\, y : \tau.\;\, x \teq
y \vee \neg \con{f}( x ) \teq \con{f}( y )$.
% where the uninterpreted type $\tau$ is
% interpreted as a finite set and $\con{f}$ as injective finite map.

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used deep inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers. A simple
example where both approaches fail is $\forall n : {\sortint}.\;\, \lite\bigl(
\con{p} \leq 0,\allowbreak\; \con{p}( n ) \teq 1,\allowbreak\; \con{p}( n ) \teq 2 * \con{p}( n - 1 ) \bigr)$.

@@

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by
$\Sigma$-interpretation $\J$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, we know $\varphi$ is of the form
$\DDD \wedge \varphi_0$. We first argue that $\I$ satisfies a $\Sigma$-formula
$\psi$ that is closed under function expansion with respect to $\DDD$, and
where $\psi$ entails $\varphi_0$.
Since $\I$ is a model of $\varphi_0$,
there exists a set of literals $\psi_0$ over the atoms occurring in $\varphi_0$
such that $\I$ satisfies $\psi_0$ and $\psi_0$ entails $\varphi_0$.
\ajr{todo}
